The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-08 manuscript/manuscript.Rnw
fix up introduction and add gestational period table

Although the morphology of the mammalian placenta has been well
characterized in a number of species \cite{Mossman.1987:vertebratefetalmembranes}, the diversity of
the molecular environment is only beginning to be understood from a
genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}. Studies of certain model species
(i.e., humans, mice, cattle) have revealed a stunning divergence in
the genetics of the placenta, especially in later stages of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
For example, each lineage has evolved expansive
placenta-specific gene families with expression patterns that can vary
in space and time \cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}. These
include independent expansions of the prolactins in mouse and cow,
pregnancy-specific glycoproteins (PSGs) in humans and mouse,
pregnancy-associated glycoproteins (PAGs) in cow, and the expansion of
the growth hormone and chorionic gonadotropin beta genes in hominoid
primates \cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}. Expression patterns of
specific genes also can vary widely between species. For example, the
expression of orthologous genes in the placenta have been found to
vary substantially between humans and mice \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
recruitment of SIGLEC6 to placental expression in humans differs from
both chimpanzees and mice \cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.

@@

One symptom
of this maternal-fetal conflict is its potential affect on human
health \cite{Hollegaard.ea2013:Parent-offspringconflictandpersistence}. 
In addition to having an invasive
hemochorial placenta, humans possess a deeply invasive trophoblast
layer that can penetrate the myometrium of the uterus 
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
 Defects in placental growth are known to be
associated with the devastating condition of preeclampsia, which can
cause hypertension, proteinuria, and has the potential to lead to
death of the mother and/or infant \DLA{(citation)}. Although cases have been
documented in chimpanzees and gorillas \DLA{(citation)}, preeclampsia has a
far higher incidence in human pregnancies with as much as 4\% of
pregnancies being affected in the United States \DLA{(citation)}.
Chimpanzees and gorillas also have deeply invasive trophoblasts,
although the progression of invasion in great apes may be slower than
humans \cite{Carter.ea2011:Evolutionofinvasiveplacentation}; however, large-scale expression
studies of great ape placentas have not been conducted. The increasing
affordability of next-generation sequencing technology has enabled the
large-scale study of specific tissues from a wide array of species to
address broader phylogenetic questions in the evolution of gene
expression \cite{Brawand.ea2011:evolutionofgeneexpression}. Here we use Illumina RNA-Seq
technology to sequence and quantify the mRNA environment of six
mammalian placentas at term, including that of the \human, \bonobo (the
sister species to the chimpanzee), \dog, \cow, \elephant, and
\opossum, representing marsupials as well as three major branches of
eutherian mammals (Euarchontoglires, Laurasiatheria, Afrotheria).
These placentas also represent a wide array of placental morphologies
as well as traits such as number and size of offspring and length of
gestation \DLA{(Table 1)}. We use this data, along with existing RNA-Seq
data from mouse \cite{Wang.ea2011:surveyfornovelimprinted}, to address the major differences
between expression in marsupial and eutherian mammals at term,
identify the evolution of expression in important placental genes such
as hormones and galectins, identify novel lineage-specific expression
patterns, and quantify human specific differences in gene expression
that may be related to obstetrical disorders such as preeclampsia.

@@

Fetal tissue was collected from the placenta of six mammalian species
(\human [human], \bonobo [bonobo], \dog [domestic dog], \cow [domestic
cow], \elephant [African savanna elephant], \opossum [gray
short-tailed opossum], \gallili [?], \carmelli [?]). Information on
specimen ID, place and time of collection, and collector is shown for
each sample in \DLA{Supplemental Table 1}. In the case of \bonobo,
\cow, and \elephant, the placenta was collected at term upon expulsion
from the mother following delivery. Placental tissue from \human was
collected upon delivery via a caesarian section. Near-term placental
tissue from \dog was collected upon the death of the mother from a car
accident. Placental tissue from \opossum was sampled at fetal stage
32, roughly one day prior to birth. Only the fetal portion of the
placenta was sampled at the maternal-fetal interface (i.e. villous
portion for \human, \bonobo, \dog, and \elephant; cotyledon for \cow;
diffuse placenta for \opossum). All tissues were collected and stored
in RNAlater (Life Technologies, Carlsbad, CA) until RNA isolation
could be performed. Tissues were homogenized in Trizol, and total RNA
was isolated following the manufacturer's instructions. RNA was
further purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in
conjunction with the RNase-Free DNase Set (Qiagen, Valencia, CA),
following the manufacturer’s protocol to remove any contaminating DNA
from the sample. Total RNA was then evaluated on a Nanodrop 1000
(Thermo Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100
(Santa Clara, CA) to confirm concentration and quality of the
isolation. RNA quality was assessed using A260/A280 and A260/A230
absorbance values.

@@

For more complete evolutionary comparisons additional sequences of
placenta in \mouse \cite{Wang.ea2011:surveyfornovelimprinted}, \human
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} \horse
\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
\DLA{citation; e-mailed Alec Archibald from the ISGC about what the
  right one should be}, and \pig
\cite{Ahanda.ea2012:Predictionofaltered3-} were obtained from the
\ac{SRA} (Supplemental \cref{tab:srr_numbers}).

@@

The
\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the transcriptome if an ortholog of that gene was expressed with
$\mathrm{FPKM} \ge 10$ in all species studied and was not a
housekeeping gene
\cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the cutoff was set
differently.
%
\textit{ANXA1} is
involved in the modulation of the inflammatory response \DLA{It does
  more stuff; read the huthcinson2011 review and fill this in}
%
Another annexin, \textit{ANXA2} 
%
\textit{TIMP3}, which may be down-regulated in pre-eclampsia
\cite{Ma.ea2014:Down-regulationofTIMP3leads} (though another study
\cite{Xiang.ea2013:PromoterhypomethylationofTIMP3} claims
up-regulation)

>>>>

Although the morphology of the mammalian placenta has been well
characterized in a number of species
\cite{Mossman.1987:vertebratefetalmembranes}, the diversity of the
molecular environment is only beginning to be understood from a
genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.
Studies of certain model species (i.e., humans, mice, cattle) have
revealed a stunning divergence in the genetics of the placenta,
especially in later stages of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
For example, each lineage has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}.
These include independent expansions of the prolactins in mouse and
cow, pregnancy-specific glycoproteins (PSGs) in humans and mouse,
pregnancy-associated glycoproteins (PAGs) in cow, and the expansion of
the growth hormone and chorionic gonadotropin beta genes in hominoid
primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between humans and mice
\cite{Su.ea2002:Large-scaleanalysisofhuman}, and the recruitment of
\textit{SIGLEC6} to placental expression in humans differs from both
chimpanzees and mice
\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.

@@

One symptom of this maternal-fetal conflict is its potential affect on
human health
\cite{Hollegaard.ea2013:Parent-offspringconflictandpersistence}. In
addition to having an invasive hemochorial placenta, humans possess a
deeply invasive trophoblast layer that can penetrate the myometrium of
the uterus \cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
Defects in placental growth are known to be associated with the
devastating condition of preeclampsia, which can cause hypertension,
proteinuria, and has the potential to lead to death of the mother
and/or infant \cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}.
Although cases have been documented in chimpanzees
\cite{Stout.ea1969:Glomerularcapillaryendothelialswelling}, and
gorillas \cite{Thornton.ea1992:Convulsionsinpregnancyin}, preeclampsia
has a far higher incidence in human pregnancies with as much as 4\% of
pregnancies being affected in the United States
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Chimpanzees
and gorillas also have deeply invasive trophoblasts, although the
progression of invasion in great apes may be slower than humans
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}; however,
large-scale expression studies of great ape placentas have not been
conducted. The increasing affordability of next-generation sequencing
technology has enabled the large-scale study of specific tissues from
a wide array of species to address broader phylogenetic questions in
the evolution of gene expression
\cite{Brawand.ea2011:evolutionofgeneexpression}. Here we use RNA-Seq
technology to sequence and quantify the mRNA environment of mammalian
placentas at term, including that of the \human, \bonobo (the sister
species to the chimpanzee), \smonkey, \mouse, \galili, \carmeli, \dog,
\cow, \elephant, \armadillo, and \opossum, representing marsupials as
well as three major branches of eutherian mammals (Euarchontoglires,
Laurasiatheria, Afrotheria). These placentas also represent a wide
array of placental morphologies as well as traits such as number and
size of offspring and length of gestation
\cref{tab:gestational_periods}. We use this data, along with existing
RNA-Seq data from \mouse \cite{Wang.ea2011:surveyfornovelimprinted},
\human \cite{Hughes.ea2015:Evaluatingintra-andinter-individual},
\horse \cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \pig
\cite{Ahanda.ea2012:Predictionofaltered3-} to address the major
differences between expression in marsupial and eutherian mammals at
term, identify the evolution of expression in important placental
genes such as hormones and galectins, identify novel lineage-specific
expression patterns, and quantify human specific differences in gene
expression that may be related to obstetrical disorders such as
preeclampsia.

\begin{table}
  \begin{tabular}{c c c c}
    \toprule
    Species                       & Gestation (days) & Litter Size  & Source \\
    \midrule
    \textit{Ateles fusciceps}     & 224 & 1 & \cite{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Bos taurus}           & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Canis familiaris}     & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Dasypus novemcinctus} & 135 & 4 & \DLA{Cite Amazing Armadillo: geography book}\\
    \textit{Equus caballus}       & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
    \textit{Loxodonta africana}   & 644 & 1 & \cite{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Homo sapiens}         & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Monodelphis domestica}& 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
    \textit{Mus musculus}         & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Nannospalax galili}   & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
    \textit{Ovis aries}           & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Pan paniscus}         & 223 & 1 & \cite{Ernest2003:lifehistorycharacteristicsofplacental} \\
    \textit{Spalax carmeli}       & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
    \textit{Sus scrofa}           & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \bottomrule
  \end{tabular}
  \caption{Gestational periods in days and average liter sizes for studied species.}
  \label{tab:gestational_periods}
\end{table}

@@

Fetal tissue was collected from the placenta of six mammalian species
(\human [human], \bonobo [bonobo], \dog [domestic dog], \cow [domestic
cow], \elephant [African savanna elephant], \opossum [gray
short-tailed opossum], \gallili [blind mole rat from Upper Galilee],
\carmelli [blind mole rat from northern Israel]). Information on
specimen ID, place and time of collection, and collector is shown for
each sample in \DLA{Supplemental Table 1}. In the case of \bonobo,
\cow, and \elephant, the placenta was collected at term upon expulsion
from the mother following delivery. Placental tissue from \human was
collected upon delivery via a caesarian section. Near-term placental
tissue from \dog was collected upon the death of the mother from a car
accident. Placental tissue from \opossum was sampled at fetal stage
32, roughly one day prior to birth. Only the fetal portion of the
placenta was sampled at the maternal-fetal interface (i.e. villous
portion for \human, \bonobo, \dog, and \elephant; cotyledon for \cow;
diffuse placenta for \opossum). All tissues were collected and stored
in RNAlater (Life Technologies, Carlsbad, CA) until RNA isolation
could be performed. Tissues were homogenized in Trizol, and total RNA
was isolated following the manufacturer's instructions. RNA was
further purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in
conjunction with the RNase-Free DNase Set (Qiagen, Valencia, CA),
following the manufacturer’s protocol to remove any contaminating DNA
from the sample. Total RNA was then evaluated on a Nanodrop 1000
(Thermo Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100
(Santa Clara, CA) to confirm concentration and quality of the
isolation. RNA quality was assessed using A260/A280 and A260/A230
absorbance values.

@@

For more complete evolutionary comparisons additional sequences of
placenta in \mouse \cite{Wang.ea2011:surveyfornovelimprinted}, \human
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} \horse
\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \pig
\cite{Ahanda.ea2012:Predictionofaltered3-} were obtained from the
\ac{SRA} (Supplemental \cref{tab:srr_numbers}).

@@

The
\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\mathrm{FPKM} \ge 10$ in all species studied and was
not a human housekeeping gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently.
%
Multiple components of the annexin complex are included in the core
transcriptome, including \textit{ANXA2}, \textit{ANXA1},
\textit{S100A11}, \textit{S100A10}. The modulation of  \DLA{It does more stuff; read
  the huthcinson2011 review and fill this in}
%
Another annexin, \textit{ANXA2} 
%
\textit{TIMP3}, which may be down-regulated in pre-eclampsia
\cite{Ma.ea2014:Down-regulationofTIMP3leads} (though another study
\cite{Xiang.ea2013:PromoterhypomethylationofTIMP3} claims
up-regulation)
